QCL: Added value of differential correlation

### 2011-11-25 inst/doc/article.Rnw
Updates to the article

<<echo=FALSE,results=hide>>=
  require(qcl)
  set.seed(200)
  data(ath.metabolites) # Arabidopsis Thaliana dataset
  cat("QCL scan\n")
  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QTL scan\n")
  qtl_result <- QTLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QCL permutation\n")
  qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,20, n.cores = 3)
  #qcl_perms <- vector("list",24)
  #idx <- 1
  #for(p in 1:24){
  #  qcl_perms[[idx]] <- read.QCLpermute("permutations", p, 300,TRUE)
  #  idx <- idx+1
  #}
@

@@

\emph{\sffamily QCL mapping}\\
  As mentioned in the introduction QCL mapping uses buffering to detect interaction partners. 
  When we split the population based on genotype at a QTL marker, only related phenotypes 
  will loosing correlation with the gene under investigation. This allows us to not only 
  only identify its interaction partners, but also means QCL mapping does not suffering from 
  (p1) many phenotypes co-localizing. Additionally it is more sensitive to detect variation 
  caused by regulatory phenotypes because variation of these regulators causes many 'downstream' 
  phenotypes to show differential correlation with the inducing phenotype, because QCL relies on 
  correlation changes as opposed to differences in phenotype mean QCL is unaffected by (p2).

@@

\emph{\sffamily Significance testing}\\
  To compare to QTL mapping we need to convert the QCLscores to LOD scores. Significance 
  of QCL were assessed by using permutations, in each round the link between genotype and 
  phenotype is broken, by redistributing at random genotypes amongst individuals not 
  allowing for duplicates. The QCL on these new individuals allows for estimation of 
  the underlying null-distribution of QCL scores. We frequently observe QCL scores in 
  real data which is higher than any QCL score obtained during permutation. When this 
  happens we use a generalized Pareto distribution (GPD) to estimate the extreme tail 
  of the distribution \cite{Knijnenburg:2009}, to also get likelyhood estimates for the 
  extreme scores observed. See figures XX, and XX showing the score distribution in 
  permuted data, Fig XX also shows a tail estimation using the GPD method.

@@

\begin{figure}[ht]
  \centering
<<permutation_scores, fig=TRUE, echo=FALSE, height=3.5>>=
op <- par(cex=0.5)
plot(c(0,0.8),c(0,400),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
hist(qcl_perms[[3]],add=T, col="yellow")
hist(qcl_perms[[2]],add=T, col="red")
hist(qcl_perms[[1]],add=T, col="green")
legend("topright",colnames(ath.metab$phenotypes)[1:3],col=c("yellow","red","green"),lwd=6)
op <- par(cex=1)
@ 
  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
\end{figure}

@@

\begin{figure}[ht]
  \centering
<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="phenotypes")
@ 
  \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:p2pmatrix}}
\end{figure}

>>>>

<<echo=FALSE,results=hide>>=
  require(qcl)
  set.seed(200)
  data(ath.metabolites) # Arabidopsis Thaliana dataset
  cat("QCL scan\n")
  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QTL scan\n")
  qtl_result <- QTLscan(ath.metab$genotypes, ath.metab$phenotypes)
  cat("QCL permutation\n")
  qcl_perms1 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 1:3, 200, n.cores = 3)
  qcl_perms2 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 4:24, 50, n.cores = 3)
  qcl_perms <- c(qcl_perms1,qcl_perms2)
  #qcl_perms <- vector("list",24)
  #idx <- 1
  #for(p in 1:24){
  #  qcl_perms[[idx]] <- read.QCLpermute("permutations", p, 300,TRUE)
  #  idx <- idx+1
  #}
@

@@

\emph{\sffamily QCL mapping}\\
  As mentioned in the introduction QCL mapping uses buffering to detect interaction partners. 
  When we split the population based on genotype at a QTL marker, only related phenotypes 
  will loose correlation with the gene under investigation. This allows us to not only 
  only identify its interaction partners, while not suffering from (p1) many phenotypes 
  co-localizing. Additionally it is more sensitive to detect variation 
  caused by regulatory phenotypes because variation of these regulators causes many 'downstream' 
  phenotypes to show differential correlation with the inducing phenotype, because QCL relies on 
  correlation changes as opposed to differences in phenotype mean QCL is unaffected by (p2).

@@

\emph{\sffamily Significance testing}\\
  To compare to QTL mapping we need to convert the QCLscores to LOD scores. Significance 
  of QCL were assessed by using permutations, in each round the link between genotype and 
  phenotype is broken, by redistributing at random genotypes amongst individuals not 
  allowing for duplicates. The QCL on these new individuals allows for estimation of 
  the underlying null-distribution of QCL scores. We frequently observe QCL scores in 
  real data which is higher than any QCL score obtained during permutation. When this 
  happens we use a generalized Pareto distribution (GPD) to estimate the extreme tail 
  of the distribution \cite{Knijnenburg:2009}, to also get likelyhood estimates for the 
  extreme scores observed. See Fig. \ref{fig:permutation_scores} showing the score 
  distribution in permuted data of 3 traits in A. Thaliana, Fig XX also shows a tail 
  estimation using the GPD method.

@@

\begin{figure}[ht]
  \centering
<<permutation_scores, fig=TRUE, echo=FALSE, height=3.0>>=
op <- par(cex=0.5)
plot(c(0,0.8),c(0,400),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
hist(qcl_perms[[3]],add=T, col="yellow")
hist(qcl_perms[[2]],add=T, col="red")
hist(qcl_perms[[1]],add=T, col="green")
legend("topright",colnames(ath.metab$phenotypes)[1:3],col=c("yellow","red","green"),lwd=6)
op <- par(cex=1)
@ 
  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
\end{figure}

@@

\begin{figure}[ht]
  \centering
<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="phenotypes")
@ 
  \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:p2pmatrix}}
\end{figure}
